The finite base property and lacking of finite axiomatisation for some subreducts of representable relation algebras

### 2020-07-04 problem-19.17/Text.tex
fix several mistakes in the definition of a rainbow algebra

\author{Daniel Rogozin}
\date{}
\title{The finite base property for some relation algebras subreducts}

@@

Given $n < \omega$, the set of atoms $\operatorname{At}(\mathfrak{A}_n)$ consists of the following elements:
\begin{itemize}
\item identity: ${\bf 1}$, an atom with no colour
\item greens: ${\bf g}_i$ for $0 \leq i \leq 2^n$
\item yellow: ${\bf y}$
\item black: ${\bf b}$
\item whites: ${\bf b}_{ij}$ for $0 \leq i \leq j \leq 2^n$
\item reds: ${\bf r}_i$ for $0 <i \leq 2^n$
\end{itemize}
We claim that every atom is self-converse ($a^{\smile} = a$). Given $x, y, z \in \mathfrak{A}_n$, a triple $(x,y,z)$ is an  inconsistent triangle if
\begin{center}
$x \land (y ; z) = y \land (z ; x) = z \land (x ; y) = 0$
\end{center}
We define the set of inconsistent triangles explicitly as follows.
\begin{itemize}
  \item $({\bf g}_i, {\bf g}_i, {\bf g}_i)$ for $0 \leq i \leq 2^n$
  \item $({\bf y}, {\bf y}, {\bf y})$ for $0 \leq i \leq 2^n$
  \item $({\bf g}_i, {\bf g}_i, {\bf w}_{kj})$ for $0 \leq i \leq 2^n$ and $0 \leq k \leq j \leq 2^n$
  \item $({\bf r}_i, {\bf r}_j, {\bf r}_k)$ unless $i + k = j$ or $i + k = j$ or $j + k = i$
  \item $({\bf g}_i, {\bf g}_{i + 1}, {\bf r})$ unless $j = 1$
  \item $({\bf g}_i, {\bf y}, {\bf w}_{jk})$ unless $i \in \{ j, k \}$
\end{itemize}
$({\bf g}_i, {\bf g}_i, {\bf w}_{kj})$ stands for ${\bf g}_i \land ({\bf g}_i ; {\bf w}_{kj}) = {\bf g}_i \land (z ; {\bf g}_i) = {\bf w}_{kj} \land ({\bf g}_i ; {\bf g}_i) = 0$, and so on.

@@

TODO: define games and networks. Take a look at \cite{hirsch1997step}.

Let $\mathcal{A}$ be a relation algebra

\begin{defin}
  A network is a complete directed finite graph with edges labelled by elements of $\mathcal{A}$. Such a graph have the following form. $N = \langle E_N, l_N \rangle$, where $E_N = U_N \times U_N$ for some finite base set and $l_n : E_N \to \operatorname{At}(\mathcal{A})$ is function mapping each edge to some atom of $A$. This function obey the following requirements:
  \begin{enumerate}
    \item $l_N(x,y) \leq {\bf 1}$ iff $x = y$
    \item $l_N(x, y) ; l_N(y,z) \geq l_N(x,z)$
  \end{enumerate}
\end{defin}
Given two networks $N = \langle E_N, l_N \rangle$ and $N^{'} = \langle E_{N^{'}}, l_{N^{'}} \rangle$, $N$ is a subnetwork of $N^{'}$ ($N \subseteq N^{'}$) if $E_N \subseteq E_{N^{'}}$ and for each $x,y \in U_N$, $l_{N^{'}}(x,y) = l_{N}(x,y)$.

@@

This lemma has been proved by Hirsch and Hodkinson here \cite{hirsch1997step}:

@@

\begin{lemma}
  Any non-trivial ultraproduct of $\{ \mathfrak{A}_n \}_{n < \omega}$ is representable. The same statement for non-trivial ultraproduct of reducts $\{ \mathfrak{S}_n \}_{n < \omega}$.
\end{lemma}

@@

\subsection{The finite algebra on finite base for ${\bf RJSOS}$ (or its failure)}

>>>>

\author{Daniel Rogozin}
\date{}
\title{The finite base property and lacking of finite axiomatisation for some subreducts of representable relation algebras}

@@

Given $n < \omega$, the set of atoms $\operatorname{At}(\mathfrak{A}_n)$ consists of the following elements:
\begin{itemize}
\item identity: ${\bf 1}$, an atom with no colour
\item greens: ${\bf g}_i$ for $0 \leq i \leq 2^n$
\item yellow: ${\bf y}$
\item black: ${\bf b}$
\item whites: ${\bf w}$, ${\bf w}_{ij}$ for $0 \leq i \leq j \leq 2^n$
\item reds: ${\bf r}_i$ for $0 <i \leq 2^n$
\end{itemize}
We claim that every atom is self-converse ($a^{\smile} = a$). Given $x, y, z \in \mathfrak{A}_n$, a triple $(x,y,z)$ is an  inconsistent triangle if
\begin{center}
$x \land (y ; z) = y \land (z ; x) = z \land (x ; y) = 0$
\end{center}
We define the set of inconsistent triangles explicitly as follows.
\begin{itemize}
  \item $({\bf 1}, x, y)$ unless $x = y$
  \item $({\bf g}_i, {\bf g}_k, {\bf g}_l)$ for $i, k, l \leq 2^n$
  \item $({\bf y}, {\bf y}, {\bf y})$
  \item $({\bf y}, {\bf y}, {\bf b})$
  \item $({\bf g}_i, {\bf g}_j, {\bf w}_{kl})$ for $i, j \leq 2^n$ and $k, j \leq 2^n$
  \item $({\bf r}_i, {\bf r}_j, {\bf r}_k)$ unless $i + k = j$ or $i + k = j$ or $j + k = i$
  \item $({\bf g}_i, {\bf g}_{i + 1}, {\bf r})$ unless $j = 1$
  \item $({\bf g}_i, {\bf y}, {\bf w}_{jk})$ unless $i \in \{ j, k \}$
\end{itemize}
$({\bf g}_i, {\bf g}_i, {\bf w}_{kj})$ stands for ${\bf g}_i \land ({\bf g}_i ; {\bf w}_{kj}) = {\bf g}_i \land ({\bf w}_{kj} ; {\bf g}_i) = {\bf w}_{kj} \land ({\bf g}_i ; {\bf g}_i) = 0$, and so on.

$C$ is the set of consistent triangles defined as

\begin{center}
  $C = \{ (a, b, c ) \: | \: a,b,c \in \operatorname{At}(\mathfrak{A}_n), (a,b,c) \text{ is consistent}\}$
\end{center}

Let us ensure briefly that $\mathfrak{A}_n$ is indeed a relation algebra defining the multiplication table for composition as in \cite{lyndon1950representation}. For that, it is enough to show associativity for atoms $a, x, y, z$ such that $a \leq (x ; y) ; z$. For that, one needs to find an atom $b$ such that $a \leq x ; b$ and $b \leq y ; z$. If $x = y$, then we put $b = {\bf 1}$. Otherwise, we try to use a white atom or a black atom for $b$ unless $x, a$ are green and $y, z$ are yellow.
Suppose $x = {\bf g}_i$, $a = {\bf g}_j$, $y = {\bf y}_k$, and $z = {\bf y}_l$, then we put

@@

Let us define networks and games according to \cite{hirsch1997step}.

\begin{defin}
  Let $\mathcal{A}$ be a relation algebra. A network is a complete directed finite graph with edges labelled by elements of $\mathcal{A}$. Such a graph have the following form. $N = \langle E_N, l_N \rangle$, where $E_N = U_N \times U_N$ for some finite base set and $l_n : E_N \to \operatorname{At}(\mathcal{A})$ is function mapping each edge to some atom of $A$. This function obey the following requirements:
  \begin{enumerate}
    \item $l_N(x,y) \leq {\bf 1}$ iff $x = y$
    \item $l_N(x, y) ; l_N(y,z) \geq l_N(x,z)$
  \end{enumerate}
\end{defin}
Given two networks $N = \langle E_N, l_N \rangle$ and $N^{'} = \langle E_{N^{'}}, l_{N^{'}} \rangle$, $N$ is a subnetwork of $N^{'}$ ($N \subseteq N^{'}$, or $N^{'}$ refines $N$) if $E_N \subseteq E_{N^{'}}$ and for each $x,y \in U_N$, $l_{N^{'}}(x,y) = l_{N}(x,y)$.

@@

This lemma has been proved by Hirsch and Hodkinson here \cite{hirsch1997step}. This lemma provide a criterion of representability for relation algebras.

@@

\begin{lemma}
  Any non-trivial ultraproduct of $\{ \mathfrak{A}_n \}_{n < \omega}$ is representable, that is, belongs to ${\bf RRA}$. The same statement for non-trivial ultraproduct of reducts $\{ \mathfrak{S}_n \}_{n < \omega}$ that belongs to $R(;, \vee)$.
\end{lemma}

@@

\subsection{The finite algebra on finite base for $R(;, \vee)$ (or its failure)}
